Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-10 paper.tex
porting my comments from pull req, also plugged in relevant cites

\raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}

@@

\raph{Answering John's question here. Short answer no. One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout. I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons. In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate. That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form. Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG. I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}

@@

\raph{To address John's comment: increasing the spin count reduces false positives; a false positive can be defined as the preceding tile posting the result late, meaning that there is some spin count for which it would proceed. However, it also delays progress in the true positive case. The deeper way to talk about this is to engage the cost model, which is done to some extent in the paragraph below. The cost of a false positive includes increased memory traffic and also compute opportunity cost of other workgroups which can be scheduled on the same multiprocessor. Indeed, with sufficient occupancy the cost of a workgroup spinning (or making scalar progress) is minimal. So optimal performance really depends on the *distribution* of spin counts. I put all this in a comment but think we can say it in the paper.}

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone~\cite{} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

>>>>

\raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}
\john{I have historically used that Lindholm reference for the GPU hardware, but for the programming model I have used Nickolls, ``Scalable parallel programming with CUDA'' (ACM Queue, Mar/Apr 2008).}

@@

\raph{Answering John's question here. Short answer no. One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout. I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons. In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate. That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form. Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG. I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.} \john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''. The reviewer, of course, doesn't actually implement this, but it totally kills the paper when a reviewer gets it into his/her head that there is another way to go about this and why didn't we study THAT way, reject, go back and incorporate my amazing idea. We need to explicitly close the door emphatically in the paper to obvious, simple, unworkable solutions; we can't give reviewers any headroom to substitute their own judgment when we know better.}

@@

\raph{To address John's comment: increasing the spin count reduces false positives; a false positive can be defined as the preceding tile posting the result late, meaning that there is some spin count for which it would proceed. However, it also delays progress in the true positive case. The deeper way to talk about this is to engage the cost model, which is done to some extent in the paragraph below. The cost of a false positive includes increased memory traffic and also compute opportunity cost of other workgroups which can be scheduled on the same multiprocessor. Indeed, with sufficient occupancy the cost of a workgroup spinning (or making scalar progress) is minimal. So optimal performance really depends on the *distribution* of spin counts. I put all this in a comment but think we can say it in the paper.}
\john{My point was not about the relationship between a spin count and false positives, it was that the draft specifically said that exposing the maximum spin count to the programmer is what minimized false positives. It's not the exposing that minimizes false positives.}

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
